---
title: "ldscore microglia TF"
author: "Katherine Tansey"
date: "21 January 2018"
output: html_document
editor_options: 
  chunk_output_type: console
---
Sanity checking the results because the enrichment and ldscore for microglia in the TFs is looking a bit messy. 

Data is bed files (chr, start, stop) with one for each TF found  in the HOMER de novo analysis (there are 20 )

# extend peaks

```{r, eval = FLASE}
module load bedtools/2.26.0

for j in Bcell BORIS BREu CEBPE DCE E2F6 Etv2 Fra1 HINFP IRF2 Mef2a RUNX Sfpi1 SP1 Spdef SPI1 Srf TFs_noAnns Usf2 Zfp281
do 
bedtools slop -i ${j}_matt_sorted.bed -g /scratch/mpnkt/reference_data/human_hg19/chromo_lengths -b 500 > ${j}_ext500.bed

done

```

# Convert bed to annot


Convert peaks to LD.annot file
*NOTE*: have to be in the baseline folder for this script to work properly....  
saved as make_annot.sh  in /scratch/mpnkt/Matthew_Hill_microglia/bin
```{r, eval = FALSE}
#!/bin/bash
#PBS -l select=1:mem=30gb
#PBS -l walltime=24:00:00
#PBS -q workq_wgp
#PBS -N make_annot
#PBS -m e
#PBS -M tanseyk@cardiff.ac.uk
#PBS -P PR341
#PBS -o OUT/
#PBS -e ERR/


module load bedtools/2.26.0

cd /scratch/mpnkt/ldsc_phase1/baseline

WORKD="/scratch/mpnkt/Matthew_Hill_microglia/ldscore_TFs"

for j in Bcell BORIS BREu CEBPE DCE E2F6 Etv2 Fra1 HINFP IRF2 Mef2a RUNX Sfpi1 SP1 Spdef SPI1 Srf TFs_noAnns Usf2 Zfp281
do     
for f in /scratch/mpnkt/ldsc_phase1/baseline/*.annot.gz
do
    NAME=$(echo ${f} | awk -F "/" '{print $6}' | awk -F ".gz" '{print $1}')
    
    echo ${NAME}
    
    gunzip ${NAME}.gz
    
    #remove header from annotation file, convert to bed and sort file
    tail -n +2 ${NAME} | awk '{print "chr"$1 "\t" $2 "\t" $2+1 "\t" $3 "\t" $4}' | sort -k1,1 -k2,2n > ${NAME}_no_head.sorted.bed
 
    #intersect with chip-seq bed file
    intersectBed -a ${NAME}_no_head.sorted.bed -b ${WORKD}/${j}_ext500.bed  -c > ${WORKD}/tmp1_${NAME}
             
    #reformat bed into LD score format
     sed 's/^chr//g' ${WORKD}/tmp1_${NAME} | awk '{print $1"\t", $2"\t", $4"\t", $5"\t", $6}' > ${WORKD}/tmp2_${NAME}
 
    echo -e "CHR\tBP\tSNP\tCM\t""SP1" | cat - ${WORKD}/tmp2_${NAME} > ${WORKD}/${j}_${NAME}
    
    wc ${NAME}
    wc ${WORKD}/${j}_${NAME}
    
    rm ${WORKD}/tmp1_${NAME}
    rm ${WORKD}/tmp2_${NAME}
    rm ${NAME}_no_head.sorted.bed
    
    gzip ${WORKD}/${j}_${NAME}
    gzip ${NAME}
    
done
done

```



# create ld files

save as : create_ld_files.sh
in : /scratch/mpnkt/Matthew_Hill_microglia/bin
```{r, eval = FALSE}
#!/bin/bash
#PBS -l select=1:mem=20gb
#PBS -l walltime=24:00:00
#PBS -q workq_wgp
#PBS -N ldscore_build
#PBS -m e
#PBS -M tanseyk@cardiff.ac.uk
#PBS -P PR341
#PBS -o OUT/
#PBS -e ERR/

module load ldsc/latest

for i in {1..22}
    do
    ldsc.py --l2 --bfile /scratch/mpnkt/ldsc_phase1/1000G_plinkfiles/1000G.mac5eur.${i} --ld-wind-cm 1  --print-snps /scratch/mpnkt/ldsc_phase1/hapmap3_snps/hm.${i}.snp --annot /scratch/mpnkt/Matthew_Hill_microglia/ldscore_TFs/Zfp281_baseline.${i}.annot.gz --out /scratch/mpnkt/Matthew_Hill_microglia/ldscore_TFs/Zfp281_baseline.${i}

done
```


# run ld score

save as : create_ld_files.sh
in : /scratch/mpnkt/Matthew_Hill_microglia/bin
```{r, eval = FALSE}
#!/bin/bash
#PBS -l select=1:mem=20gb
#PBS -l walltime=24:00:00
#PBS -q workq_wgp
#PBS -N ldscore 
#PBS -m e
#PBS -M tanseyk@cardiff.ac.uk
#PBS -P PR341

module load ldsc/latest

for j in Bcell BORIS BREu CEBPE DCE E2F6 Etv2 Fra1 HINFP IRF2 Mef2a RUNX Sfpi1 SP1 Spdef SPI1 Srf TFs_noAnns Usf2 Zfp281
do 

ldsc.py  --h2  /scratch/mpnkt/Matthew_Hill_microglia/SP1_ldscore/IGAP_stage_1.sumstats.gz --ref-ld-chr /scratch/mpnkt/ldsc_phase1/baseline/baseline.,/scratch/mpnkt/Matthew_Hill_microglia/ldscore_TFs/${j}_baseline. --w-ld-chr /scratch/mpnkt/ldsc_phase1/weights_hm3_no_hla/weights. --overlap-annot --frqfile-chr /scratch/mpnkt/ldsc_phase1/1000G_frq/1000G.mac5eur.  --out /scratch/mpnkt/Matthew_Hill_microglia/ldscore_TFs/${j}_500ext --print-coefficients

done 
```

# R analysis  


```{r, eval=FALSE}

### do this for every folder to one results file. 
for i in   Bcell BORIS BREu CEBPE DCE E2F6 Etv2 Fra1 HINFP IRF2 Mef2a RUNX Sfpi1 SP1 Spdef SPI1 Srf TFs_noAnns Usf2 Zfp281
do
    foo="$(grep _1 ${i}_500ext.results) "
    echo $i "$foo" >> AD_TF_ldscore
done
```


```{r}
library(stringr)
library(reshape2)
library(ggplot2)
library(dplyr)
library(knitr)
library(grid)

# Load the names of all the cell types in
cell_types <- read.table("/Users/katherine/Dropbox/ldsc/cell_types.txt", h=T)
cell_types$Cell_Marks = paste(cell_types$cell_type, cell_types$mark, sep="_")
names(cell_types)[names(cell_types)=="file_number"] <- "NUMBER"
cell_types_1 <- cell_types[,c(1,4)]
```

## AD -- TF

```{r}
AD_TF <- read.table("/Users/katherine/Documents/Data_Clinic_Work/Matthew_Hill/microgilia/ldscore_TF/AD_TF_ldscore")

names(AD_TF) <- c("TF", "Category", "Prop._SNPs", "Prop._h2", "Prop._h2_std_error", "Enrichment", "Enrichment_std_error", "Enrichment_p", "Coefficient", "Coefficient_std_error", "Coefficient_zscore")

AD_TF$pvalue <- ifelse(AD_TF$Coefficient_zscore > 0,
                              2*pnorm(-abs(AD_TF$Coefficient_zscore)),
                               1-(2*pnorm(-abs(AD_TF$Coefficient_zscore))))


AD_TF$neglog10p <- -log10(AD_TF$pvalue)

sig = -log10(0.05/nrow(AD_TF))

### multiple correction line
print(sig)
```


Blue colored bars are the cell marks of interest.  
```{r, echo=FALSE, fig.width=7, fig.height=20}
ggplot(AD_TF,  aes(TF, neglog10p)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    theme_bw() +
    theme(axis.text.x = element_text(vjust=1, hjust=1, size=10),
          axis.text.y = element_text(size=7),
          legend.position="none")+
    xlab("Category") +
    geom_hline(yintercept=sig, color="red") +
    ylab("-log10 P-value")
```

```{r, echo=FALSE}
### make table for top results
kable(AD_TF %>%
    select(label, Prop._SNPs, Prop._h2, Coefficient_z.score, pvalue, neglog10p) %>%
    arrange(pvalue))
```

